---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# bestNormalize: Flexibly calculate the best normalizing transformation for a vector

The `bestNormalize` R package was designed to help find a normalizing transformation for a vector. There are many techniques that have been developed in this aim, however each has been subject to their own strengths/weaknesses, and it is unclear on how to decide which will work best until the data is oberved. This package will look at a range of possible transformations and return the best one, i.e. the one that makes it look the *most* normal. 

This package also introduces a new normalization technique, Ordered Quantile normalization (`orderNorm()`), which transforms the data based off of a rank mapping to the normal distribution, which allows us to *guarantee* normally distributed transformed data (if ties are not present). 

## Installation

You can install bestNormalize from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("petersonR/bestNormalize")
```

## Example

In this example, we generate 1000 draws from a gamma distribution, and normalize them:

```{r example}
library(bestNormalize)
set.seed(100)
x <- rgamma(1000, 1, 1)

# Estimate best transformation
BN_obj <- bestNormalize(x)
BN_obj

# Perform transformation
gx <- predict(BN_obj)

# Perform reverse transformation
x2 <- predict(BN_obj, newdata = gx, inverse = TRUE)

# Prove the transformation is 1:1
all.equal(x2, x)

```
